Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-06-24 parallel-tractability-J-v9/parallelTractabilityOntMat.tex
revision of the constant time claims

\begin{example}\label{exp:dllite}
Given a DL-Lite ontology $\mathcal{O}_{\text{ex}_\ref{exp:dllite}}$
where the TBox and RBox contain the following axioms:
$A\sqsubseteq B_1$, $A\sqsubseteq B_2$, $B_1\sqcap B_2\sqsubseteq\bot$,
$\exists R.B_2\sqsubseteq\bot$, $Q\sqsubseteq S^-$, $\exists Q.\top\sqsubseteq A$ and $S\sqsubseteq R$;
its ABox contains an assertion $Q(a,b)$.
We denote the corresponding datalog program of $\mathcal{O}_{\text{ex}_\ref{exp:dllite}}$ by $P_{\text{ex}_\ref{exp:dllite}}=\langle R, \textbf{I}\rangle$,
where $R$ contains the rules that are transformed from the above axioms; $\textbf{I}$ contains $Q(a,b)$
as the only fact.
The unique materialization graph of $P_{\text{ex}_\ref{exp:dllite}}$ is denoted by
$\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$ (see
Figure~\ref{fig:ex2}).
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{fig-dllite.pdf}
\caption{The materialization graph of $\mathcal{O}_{\text{ex}_\ref{exp:dllite}}$}
\label{fig:ex2}
\end{center}
\end{figure}

Consider performing Algorithm~$\mathsf{A}_{\text{opt}}$ on the ontology $\mathcal{O}_{\text{ex}_\ref{exp:dllite}}$ in Example~\ref{exp:dllite}.
(\uppercase\expandafter{\romannumeral1}) First, Algorithm~$\mathsf{A}_{\text{opt}}$ adds all ABox assertions (only $Q(a,b)$ here)
to the initially empty graph $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$ (\ref{alg3:addFacts} of Algorithm~$\mathsf{A}_{\text{opt}}$).
(\uppercase\expandafter{\romannumeral2}) In the first iteration of
\ref{alg3:updateG}, Algorithm~$\mathsf{A}_{\text{opt}}$ checks that all of the nodes
$A(a)$, $S(b,a)$, $B_1(a)$, $B_2(a)$ and $R(b,a)$ have corresponding SWD paths
starting from $Q(a,b)$. Thus, Algorithm~$\mathsf{A}_{\text{opt}}$
adds these nodes to $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$ immediately. We take an example of node $B_1(a)$,
for which an SWD path $(Q(a,b),A(a),B_1(a))$ exists.
When updating $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$ by adding $B_1(a)$ to it,
Algorithm~$\mathsf{A}_{\text{opt}}$ first checks whether the parent node $A(a)$ of $B_1(a)$
has been in $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$; if $A(a)$ is already in $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$, $B_1(a)$
is added to $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$ by creating an edge pointing from $A(a)$ to $B_1(a)$;
if $A(a)$ has not been added to $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$,
Algorithm~$\mathsf{A}_{\text{opt}}$ adds $A(a)$ ($B_1(a)$, respectively)
to $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$ and creates an edge pointing from
$Q(a,b)$ to $A(a)$ (an edge pointing from $A(a)$ to $B_1(a)$, respectively).\footnote{These two cases may happen simultaneously,
since node $A(a)$ and node $B_1(a)$ are being processed in parallel.} The other nodes
$A(a)$, $S(b,a)$, $B_2(a)$ and $R(b,a)$ are processed similarly.
(\uppercase\expandafter{\romannumeral3})
The two nodes $\bot(a)$ and $\bot(b)$
have no SWD path in the first iteration of \ref{alg3:updateG}.
They are left to be processed in the second iteration.
(\uppercase\expandafter{\romannumeral4})
Finally, Algorithm~$\mathsf{A}_{\text{opt}}$ finishes constructing $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$
after two iterations (\ref{alg3:halt}).

@@

The above lemma guarantees that all atoms of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$
have to be added to the constructed materialization graph in the first iteration of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{\text{opt}}$.
For each atom of the form $\bot(x)$, there may not exist an SWD path in the first iteration
of \ref{alg3:updateG}, since its derivability depends on its two parent nodes
(see nodes $\bot(a)$ and $\bot(b)$ in $\mathcal{G}_{\text{ex}_\ref{exp:dllite}}$ of
Example~\ref{exp:dllite}). On the other hand, according to the syntax of DL-Lite,
$\bot$ does not occur on the left-hand side of any axiom. In other words,
an atom of the form $\bot(x)$ cannot be the parent of any other node.
This allows for adding all atoms of the form $\bot(x)$ to the constructed materialization graph in at most two iterations of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{\text{opt}}$.
Based on the above discussion, for any DL-Lite ontology $\mathcal{O}$,
there always exists a poly-logarithmically bounded function $\psi$ such that
Algorithm~$\mathsf{A}_{\text{opt}}^{\psi}$ can handle the materialization of $\mathcal{O}$;
more precisely, we can set that $\psi=2$. This result is consistent with
that by \citet{CalvaneseGLLR07}.
Formally, we use $\mathcal{D}_{\textit{dl\_lite}}$
to denote the set of all DL-Lite$_{\text{core}}$ and DL-Lite$_\mathcal{R}$ ontologies and give the following theorem.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{fig-pathtwist.pdf}
\caption{The materialization graph of $\mathcal{O}_{\text{ex}_\ref{exp:dhl}}$
  (left) and of $\mathcal{O}_{\text{ex}_\ref{exp:simpleC}}$ (right)}
\label{fig:ex3_4}
\end{center}
\end{figure}

\begin{example}\label{exp:dhl}
Given a DHL ontology $\mathcal{O}_{\text{ex}_\ref{exp:dhl}}$ where the TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $\exists R.A\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), R(a_i,a_{i-1}), A(a_1)\}$
for $2\leq i\leq k$ and $k$ an integer greater than $2$.
We denote the corresponding datalog program of $\mathcal{O}_{\text{ex}_\ref{exp:dhl}}$ by $P_{\text{ex}_\ref{exp:dhl}}=\langle R, \textbf{I}\rangle$,
where $R$ contains three rules: `$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$R(x,y),A(y)\rightarrow B_2(x)$'.
The materialization graph of $P_{\text{ex}_\ref{exp:dhl}}$ constructed by
$\mathsf{A}_{\text{opt}}$ is denoted by $\mathcal{G}_{\text{ex}_\ref{exp:dhl}}$
(see Figure~\ref{fig:ex3_4}~(left)).
\end{example}

One can check that $\mathcal{G}_{\text{ex}_\ref{exp:dhl}}$ is the unique materialization graph of $P_{\text{ex}_\ref{exp:dhl}}$.
Observe that there exists a path (e.g., $A(a_1),B_1(a_2),A(a_2),\ldots,A(a_k)$) between $A(a_1)$ and $A(a_k)$.
When performing Algorithm~$\mathsf{A}_{\text{opt}}$ on $P_{\text{ex}_\ref{exp:dhl}}$, it can be checked that
each node of the form $A(a_i)$ (filled with black color,
and $2\leq i\leq k$) has no SWD path until the $i^{th}$ iteration;
in the $i^{th}$ iteration, the parent nodes of $A(a_i)$ have been added to $\mathcal{G}_{\text{ex}_\ref{exp:dhl}}$,
and, $A(a_i)$ can also be added to $\mathcal{G}_{\text{ex}_\ref{exp:dhl}}$.
Similar to the datalog program class $\mathbb{P}_{\text{ex}_\ref{exp:mg}}$, we can also get a
datalog program class $\mathbb{P}_{\text{ex}_\ref{exp:dhl}}$ for the ontology $\mathcal{O}_{\text{ex}_\ref{exp:dhl}}$
when $k$ is a variable. Based on the above analysis, Algorithm~$\mathsf{A}_{\text{opt}}$ cannot complete the materialization
for the datalog programs of $\mathbb{P}_{\text{ex}_\ref{exp:dhl}}$ in a poly-logarithmic
number of iterations.
The intuitive reason is that at least two paths exist
from $A(a_1)$ to $A(a_k)$. These paths \emph{twist} mutually and share the same joint nodes (see the black nodes).
This invalidates the optimization used in Algorithm~$\mathsf{A}_{\text{opt}}$.
That is, for each node $A(a_i)$, $2\leq i\leq k$, until its parents ($B_1(a_i)$ and $B_2(a_i)$) are added
to $\mathcal{G}_{\text{ex}_\ref{exp:dhl}}$, there would not exist an available SWD path for $A(a_i)$.
We use the term `\emph{path twisting}' to refer to such cases.

@@

\begin{example}\label{exp:simpleC}
Consider an ontology $\mathcal{O}_{\text{ex}_\ref{exp:simpleC}}$ where the TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $B_3\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), B_3(a_i), A(a_1)\}$
for $2\leq i\leq k$ and $k$ an integer greater than $2$.
We denote the corresponding datalog program by $P_{\text{ex}_\ref{exp:simpleC}}$ where the rule set contains:
`$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$B_3(x)\rightarrow B_2(x)$'.
$P_{\text{ex}_\ref{exp:simpleC}}$ has a unique materialization graph denoted by $\mathcal{G}_{\text{ex}_\ref{exp:simpleC}}$ (see Figure~\ref{fig:ex3_4}~(right)).
\end{example}

In the above example, for the axiom $B_1\sqcap B_2\sqsubseteq A$, all derived atoms
of the form $B_2(x)$ must not be child nodes of an atom $A(y)$ for some $y$.
This ensures that only one path exists between each two nodes among $A(a_2),\ldots,A(a_k)$.
Further, when constructing $\mathcal{G}_{\text{ex}_\ref{exp:simpleC}}$, Algorithm~$\mathsf{A}_{\text{opt}}$ can terminate after
two iterations of \ref{alg3:updateG}. Specifically, in the first iteration, Algorithm~$\mathsf{A}_{\text{opt}}$
adds all of the nodes $B_3(a_i)$ and $B_2(a_i)$, $2\leq i\leq k$, to $\mathcal{G}_{\text{ex}_\ref{exp:simpleC}}$
since they have corresponding SWD paths;
after that, all other nodes
can be added to $\mathcal{G}_{\text{ex}_\ref{exp:simpleC}}$ in the second iteration (because each
node has an SWD path).
Motivated by this example, we consider restricting the usage of
the axioms $B_1\sqcap B_2\sqsubseteq A$
such that all atoms of the form $B_1(x)$ or $B_2(x)$ cannot be generated by an atom $A(y)$
for some $y$.
To this end, we first define \emph{simple concepts} as follows:

@@

One can check that the materialization graph $\mathcal{G}_{\text{ex}_{\ref{exp:complexRIA}}}$ has the same shape as that
of $\mathcal{G}_{\text{ex}_\ref{exp:dhl}}$ in Figure~\ref{fig:ex3_4}.
A twisted path exists in $\mathcal{G}_{\text{ex}_{\ref{exp:complexRIA}}}$ involving
$R(a_i,a_i)$, $2\leq i\leq k$, as the joint nodes.
Further, all the roles $R_1$, $R_2$, $R_3$, $R_4$ and $R$ in this example are non-transitive roles.

@@

\begin{example}
For the ontology $\mathcal{O}_{\text{ex}_{\ref{exp:complexRIA}}}$ in Example~\ref{exp:complexRIA}, all of the roles
$R_1$, $R_2$ and $R$ are non-simple roles. Thus, $\mathcal{O}_{\text{ex}_{\ref{exp:complexRIA}}}$ does not follow
the simple-role restriction because of $R_1\circ R_2\sqsubseteq R$.
Consider the ontology $\mathcal{O}_{\text{ex}_\ref{exp:mg}}$
in Example~\ref{exp:mg} again. The role $R$ is a non-simple role, while $S$ is a simple role.
Thus, $\mathcal{O}_{\text{ex}_\ref{exp:mg}}$ follows the simple-role restriction. All the implicit nodes
in $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ have corresponding SWD paths in the first iteration.
Thus, `path twisting' cannot occur when materializing $\mathcal{O}_{\text{ex}_\ref{exp:mg}}$ by Algorithm~$\mathsf{A}_{\text{opt}}$.
\end{example}

>>>>

\begin{example}\label{exp:dllite}
Given a DL-Lite ontology $\mathcal{O}_{\text{ex}_{\ref{exp:dllite}}}$
where the TBox and RBox contain the following axioms:
$A\sqsubseteq B_1$, $A\sqsubseteq B_2$, $B_1\sqcap B_2\sqsubseteq\bot$,
$\exists R.B_2\sqsubseteq\bot$, $Q\sqsubseteq S^-$, $\exists Q.\top\sqsubseteq A$ and $S\sqsubseteq R$;
its ABox contains an assertion $Q(a,b)$.
We denote the corresponding datalog program of $\mathcal{O}_{\text{ex}_{\ref{exp:dllite}}}$ by $P_{\text{ex}_{\ref{exp:dllite}}}=\langle R, \textbf{I}\rangle$,
where $R$ contains the rules that are transformed from the above axioms; $\textbf{I}$ contains $Q(a,b)$
as the only fact.
The unique materialization graph of $P_{\text{ex}_{\ref{exp:dllite}}}$ is denoted by
$\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$ (see
Figure~\ref{fig:ex2}).
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{fig-dllite.pdf}
\caption{The materialization graph of $\mathcal{O}_{\text{ex}_{\ref{exp:dllite}}}$}
\label{fig:ex2}
\end{center}
\end{figure}

Consider performing Algorithm~$\mathsf{A}_{\text{opt}}$ on the ontology $\mathcal{O}_{\text{ex}_{\ref{exp:dllite}}}$ in Example~\ref{exp:dllite}.
(\uppercase\expandafter{\romannumeral1}) First, Algorithm~$\mathsf{A}_{\text{opt}}$ adds all ABox assertions (only $Q(a,b)$ here)
to the initially empty graph $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$ (\ref{alg3:addFacts} of Algorithm~$\mathsf{A}_{\text{opt}}$).
(\uppercase\expandafter{\romannumeral2}) In the first iteration of
\ref{alg3:updateG}, Algorithm~$\mathsf{A}_{\text{opt}}$ checks that all of the nodes
$A(a)$, $S(b,a)$, $B_1(a)$, $B_2(a)$ and $R(b,a)$ have corresponding SWD paths
starting from $Q(a,b)$. Thus, Algorithm~$\mathsf{A}_{\text{opt}}$
adds these nodes to $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$ immediately. We take an example of node $B_1(a)$,
for which an SWD path $(Q(a,b),A(a),B_1(a))$ exists.
When updating $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$ by adding $B_1(a)$ to it,
Algorithm~$\mathsf{A}_{\text{opt}}$ first checks whether the parent node $A(a)$ of $B_1(a)$
has been in $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$; if $A(a)$ is already in $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$, $B_1(a)$
is added to $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$ by creating an edge pointing from $A(a)$ to $B_1(a)$;
if $A(a)$ has not been added to $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$,
Algorithm~$\mathsf{A}_{\text{opt}}$ adds $A(a)$ ($B_1(a)$, respectively)
to $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$ and creates an edge pointing from
$Q(a,b)$ to $A(a)$ (an edge pointing from $A(a)$ to $B_1(a)$, respectively).\footnote{These two cases may happen simultaneously,
since node $A(a)$ and node $B_1(a)$ are being processed in parallel.} The other nodes
$A(a)$, $S(b,a)$, $B_2(a)$ and $R(b,a)$ are processed similarly.
(\uppercase\expandafter{\romannumeral3})
The two nodes $\bot(a)$ and $\bot(b)$
have no SWD path in the first iteration of \ref{alg3:updateG}.
They are left to be processed in the second iteration.
(\uppercase\expandafter{\romannumeral4})
Finally, Algorithm~$\mathsf{A}_{\text{opt}}$ finishes constructing $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$
after two iterations (\ref{alg3:halt}).

@@

The above lemma guarantees that all atoms of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$
have to be added to the constructed materialization graph in the first iteration of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{\text{opt}}$.
For each atom of the form $\bot(x)$, there may not exist an SWD path in the first iteration
of \ref{alg3:updateG}, since its derivability depends on its two parent nodes
(see nodes $\bot(a)$ and $\bot(b)$ in $\mathcal{G}_{\text{ex}_{\ref{exp:dllite}}}$ of
Example~\ref{exp:dllite}). On the other hand, according to the syntax of DL-Lite,
$\bot$ does not occur on the left-hand side of any axiom. In other words,
an atom of the form $\bot(x)$ cannot be the parent of any other node.
This allows for adding all atoms of the form $\bot(x)$ to the constructed materialization graph in at most two iterations of
\ref{alg3:updateG} by applying Algorithm~$\mathsf{A}_{\text{opt}}$.
Based on the above discussion, for any DL-Lite ontology $\mathcal{O}$,
there always exists a poly-logarithmically bounded function $\psi$ such that
Algorithm~$\mathsf{A}_{\text{opt}}^{\psi}$ can handle the materialization of $\mathcal{O}$;
more precisely, we can set that $\psi=2$. This result is consistent with
that by \citet{CalvaneseGLLR07}.
Formally, we use $\mathcal{D}_{\textit{dl\_lite}}$
to denote the set of all DL-Lite$_{\text{core}}$ and DL-Lite$_\mathcal{R}$ ontologies and give the following theorem.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{fig-pathtwist.pdf}
\caption{The materialization graph of $\mathcal{O}_{\text{ex}_{\ref{exp:dhl}}}$
  (left) and of $\mathcal{O}_{\text{ex}_{\ref{exp:simpleC}}}$ (right)}
\label{fig:ex3_4}
\end{center}
\end{figure}

\begin{example}\label{exp:dhl}
Given a DHL ontology $\mathcal{O}_{\text{ex}_{\ref{exp:dhl}}}$ where the TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $\exists R.A\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), R(a_i,a_{i-1}), A(a_1)\}$
for $2\leq i\leq k$ and $k$ an integer greater than $2$.
We denote the corresponding datalog program of $\mathcal{O}_{\text{ex}_{\ref{exp:dhl}}}$ by $P_{\text{ex}_{\ref{exp:dhl}}}=\langle R, \textbf{I}\rangle$,
where $R$ contains three rules: `$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$R(x,y),A(y)\rightarrow B_2(x)$'.
The materialization graph of $P_{\text{ex}_{\ref{exp:dhl}}}$ constructed by
$\mathsf{A}_{\text{opt}}$ is denoted by $\mathcal{G}_{\text{ex}_{\ref{exp:dhl}}}$
(see Figure~\ref{fig:ex3_4}~(left)).
\end{example}

One can check that $\mathcal{G}_{\text{ex}_{\ref{exp:dhl}}}$ is the unique materialization graph of $P_{\text{ex}_{\ref{exp:dhl}}}$.
Observe that there exists a path (e.g., $A(a_1),B_1(a_2),A(a_2),\ldots,A(a_k)$) between $A(a_1)$ and $A(a_k)$.
When performing Algorithm~$\mathsf{A}_{\text{opt}}$ on $P_{\text{ex}_{\ref{exp:dhl}}}$, it can be checked that
each node of the form $A(a_i)$ (filled with black color,
and $2\leq i\leq k$) has no SWD path until the $i^{th}$ iteration;
in the $i^{th}$ iteration, the parent nodes of $A(a_i)$ have been added to $\mathcal{G}_{\text{ex}_{\ref{exp:dhl}}}$,
and, $A(a_i)$ can also be added to $\mathcal{G}_{\text{ex}_{\ref{exp:dhl}}}$.
Similar to the datalog program class $\mathbb{P}_{\text{ex}_{\ref{exp:mg}}}$, we can also get a
datalog program class $\mathbb{P}_{\text{ex}_{\ref{exp:dhl}}}$ for the ontology $\mathcal{O}_{\text{ex}_{\ref{exp:dhl}}}$
when $k$ is a variable. Based on the above analysis, Algorithm~$\mathsf{A}_{\text{opt}}$ cannot complete the materialization
for the datalog programs of $\mathbb{P}_{\text{ex}_{\ref{exp:dhl}}}$ in a poly-logarithmic
number of iterations.
The intuitive reason is that at least two paths exist
from $A(a_1)$ to $A(a_k)$. These paths \emph{twist} mutually and share the same joint nodes (see the black nodes).
This invalidates the optimization used in Algorithm~$\mathsf{A}_{\text{opt}}$.
That is, for each node $A(a_i)$, $2\leq i\leq k$, until its parents ($B_1(a_i)$ and $B_2(a_i)$) are added
to $\mathcal{G}_{\text{ex}_{\ref{exp:dhl}}}$, there would not exist an available SWD path for $A(a_i)$.
We use the term `\emph{path twisting}' to refer to such cases.

@@

\begin{example}\label{exp:simpleC}
Consider an ontology $\mathcal{O}_{\text{ex}_{\ref{exp:simpleC}}}$ where the TBox contains three axioms:
$B_1\sqcap B_2\sqsubseteq A$, $\exists S.A\sqsubseteq B_1$ and $B_3\sqsubseteq B_2$;
the ABox is $\{S(a_i,a_{i-1}), B_3(a_i), A(a_1)\}$
for $2\leq i\leq k$ and $k$ an integer greater than $2$.
We denote the corresponding datalog program by $P_{\text{ex}_{\ref{exp:simpleC}}}$ where the rule set contains:
`$B_1(x),B_2(x)\rightarrow A(x)$',`$S(x,y),A(y)\rightarrow B_1(x)$' and `$B_3(x)\rightarrow B_2(x)$'.
$P_{\text{ex}_{\ref{exp:simpleC}}}$ has a unique materialization graph denoted by $\mathcal{G}_{\text{ex}_{\ref{exp:simpleC}}}$ (see Figure~\ref{fig:ex3_4}~(right)).
\end{example}

In the above example, for the axiom $B_1\sqcap B_2\sqsubseteq A$, all derived atoms
of the form $B_2(x)$ must not be child nodes of an atom $A(y)$ for some $y$.
This ensures that only one path exists between each two nodes among $A(a_2),\ldots,A(a_k)$.
Further, when constructing $\mathcal{G}_{\text{ex}_{\ref{exp:simpleC}}}$, Algorithm~$\mathsf{A}_{\text{opt}}$ can terminate after
two iterations of \ref{alg3:updateG}. Specifically, in the first iteration, Algorithm~$\mathsf{A}_{\text{opt}}$
adds all of the nodes $B_3(a_i)$ and $B_2(a_i)$, $2\leq i\leq k$, to $\mathcal{G}_{\text{ex}_{\ref{exp:simpleC}}}$
since they have corresponding SWD paths;
after that, all other nodes
can be added to $\mathcal{G}_{\text{ex}_{\ref{exp:simpleC}}}$ in the second iteration (because each
node has an SWD path).
Motivated by this example, we consider restricting the usage of
the axioms $B_1\sqcap B_2\sqsubseteq A$
such that all atoms of the form $B_1(x)$ or $B_2(x)$ cannot be generated by an atom $A(y)$
for some $y$.
To this end, we first define \emph{simple concepts} as follows:

@@

One can check that the materialization graph $\mathcal{G}_{\text{ex}_{\ref{exp:complexRIA}}}$ has the same shape as that
of $\mathcal{G}_{\text{ex}_{\ref{exp:dhl}}}$ in Figure~\ref{fig:ex3_4}.
A twisted path exists in $\mathcal{G}_{\text{ex}_{\ref{exp:complexRIA}}}$ involving
$R(a_i,a_i)$, $2\leq i\leq k$, as the joint nodes.
Further, all the roles $R_1$, $R_2$, $R_3$, $R_4$ and $R$ in this example are non-transitive roles.

@@

\begin{example}
For the ontology $\mathcal{O}_{\text{ex}_{\ref{exp:complexRIA}}}$ in Example~\ref{exp:complexRIA}, all of the roles
$R_1$, $R_2$ and $R$ are non-simple roles. Thus, $\mathcal{O}_{\text{ex}_{\ref{exp:complexRIA}}}$ does not follow
the simple-role restriction because of $R_1\circ R_2\sqsubseteq R$.
Consider the ontology $\mathcal{O}_{\text{ex}_{\ref{exp:mg}}}$
in Example~\ref{exp:mg} again. The role $R$ is a non-simple role, while $S$ is a simple role.
Thus, $\mathcal{O}_{\text{ex}_{\ref{exp:mg}}}$ follows the simple-role restriction. All the implicit nodes
in $\mathcal{G}_{\text{ex}_{\ref{exp:mg}}}$ have corresponding SWD paths in the first iteration.
Thus, `path twisting' cannot occur when materializing $\mathcal{O}_{\text{ex}_{\ref{exp:mg}}}$ by Algorithm~$\mathsf{A}_{\text{opt}}$.
\end{example}

### 2018-06-24 parallel-tractability-J-v9/response2.tex
revision of the constant time claims

\begin{quote}
  In Algorithm $\mathsf{A}_\text{bst}$, the reason why Step~2 costs
  constant time is still unclear. The paragraph below Lemma~1 does not
  provide any explanation. The authors pointed out two references in
  their response, but I could not find any clue in those articles (one
  of which is just a two page summary \textcolor{red}{I can also not
    see why this reference is of any help. Maybe we should comment on
    this? }). \textcolor{red}{For the second reference
    \cite{Raymond95}, I can also not see where it talks about matrix
    multiplication? Does this follow from what is written below
    Def.~3.2.10?} Again, even under the assumption that there is an
  one-to-one correspondence between processors and rule instances, for
  each processor and the corresponding rule $B_1, \ldots, B_n \to H$,
  the complexity of checking the applicability of
  $B_1, \ldots, B_n \to H$ in $G$ and at the same time updating $G$
  should depend on the size of $G$.
\end{quote}

One issue might have been our use of the two terms `rule instance'
and `ground instantiation of a rule', which denote the same thing (but
this was maybe not clear in the paper).
We now unify the terms and only use `rule instantiation', which is introduced
in Sec.~2.1. A rule instantiation is different from a rule.
A rule instantiation has no variable while a rule can have variables.
Checking the applicability of a rule depends on the size of the
processed graph. However, checking the applicability of a rule
instantiation only depends on the derivability of the ground atoms
in this rule instantiation. Suppose that a processor can access the
location of any atom in constant time. It can also check the applicability of
a rule \emph{instantiation} in constant time.

@@

1) Suppose that there is a one-to-one correspondence between
processors and rule instantiations, and each processor can access the
location of any atom \textcolor{red}{within a rule? atom here means
  derived facts?} in constant time
(given as the parallel assumption in Section~3.3). Since a processor
$p$ corresponds to a rule instantiation, e.g.,
$B_1, \ldots, B_n \to H$, processor $p$ knowns the relationship among
$B_1, \ldots, B_n$ and $H$. Further, processor $p$ knows the locations
of the nodes $B_1, \ldots, B_n$ and $H$ \textcolor{red}{within the
  materialization graph? how?  by checking (in constant time) the
  derived facts?}. Thus, it could check the
derivability of $B_1, \ldots, B_n$ \textcolor{red}{Check the
  derivability or whether the (ground) atoms have been derived?}, and add an edge from $B_i$ to $H$
($1 \leq i \leq n$) in constant time instead of searching the whole
graph $G$.

2) How to establish such a one-to-one correspondence between processors
and rule instantiations and the constant-time access to each atom for processors?

We should also introduce the notion of Boolean circuits, which are 
circuits that consist of logic gates, e.g., AND and OR gates (see
\cite[Section~2.3]{Raymond95}).  A materialization graph can also be
seen as a Boolean circuit by treating the nodes as AND gates.\\
%
First, the evaluation of a Datalog program is equivalent to the evaluation
of its corresponding Boolean circuit in terms of computational
complexity \cite{DMRT14a}. \\
%
Second, a Boolean circuit can be simulated by PRAM based on an
algorithm called parallel prefix computation \cite{LaFi80a}.  Further,
the complexity of simulation is lower than that of the evaluation of
the Boolean circuit \cite[Lemma~2.4.1]{Raymond95}.\\
%
Third, during the above simulation procedure, a one-to-one
correspondence between each gate in the Boolean circuit and the
processors in PRAM and the constant-time access are established.\\
%
In our case, the one-to-one correspondence between processors and rule
instantiations and the constant-time access to each node can also be
established by PRAM similarly to the above simulation procedure.

\textcolor{red}{We have added the above argumentation also to the
  paper. (needed?, TBD) \emph{Rephrase the relevant paragraphs in a brief way?}}

@@

\textcolor{red}{The argumentation for this claim is similar to the one
  given above. (?) Is it? Can we simply reply like that?}

>>>>

\begin{quote}
  In Algorithm $\mathsf{A}_\text{bst}$, the reason why Step~2 costs
  constant time is still unclear. The paragraph below Lemma~1 does not
  provide any explanation. The authors pointed out two references in
  their response, but I could not find any clue in those articles (one
  of which is just a two page summary. Again, even under the assumption that there is an
  one-to-one correspondence between processors and rule instances, for
  each processor and the corresponding rule $B_1, \ldots, B_n \to H$,
  the complexity of checking the applicability of
  $B_1, \ldots, B_n \to H$ in $G$ and at the same time updating $G$
  should depend on the size of $G$.
\end{quote}

One issue might have been our use of the two terms `rule instance'
and `ground instantiation of a rule', which denote the same thing (but
this was maybe not clear in the paper).
We now unify the terms and only use `rule instantiation', which is introduced
in Sec.~2.1. A rule instantiation is different from a rule.
A rule instantiation has no variable while a rule can have variables.
Checking the applicability of a rule depends on the size of the
processed graph. However, checking the applicability of a rule
instantiation only depends on the derivability of the ground atoms
in this rule instantiation. Suppose that a processor can access the
locations of the atoms in the corresponding rule instantiation in constant time.
It can also check the applicability of a rule \emph{instantiation} in constant time.

@@

1) Suppose that there is a one-to-one correspondence between
processors and rule instantiations, and each processor can access the
locations of the atoms within the rule instantiation in constant time
(given as the parallel condition in Section~3.3). Since a processor
$p$ corresponds to a rule instantiation, e.g.,
$B_1, \ldots, B_n \to H$, processor $p$ knows the relationship among
$B_1, \ldots, B_n$ and $H$. Further, processor $p$ knows the locations
of the nodes $B_1, \ldots, B_n$ and $H$ which are in the rule instantiation $B_1, \ldots, B_n \to H$.
Thus, it could check whether the nodes $B_1, \ldots, B_n$ have been derived,
and add an edge from $B_i$ to $H$ ($1 \leq i \leq n$) in constant time instead of searching the whole
graph $G$.

2) How to establish such a one-to-one correspondence between processors
and rule instantiations and the constant-time access to atoms for processors?

We should also introduce the notion of Boolean circuits, which are
circuits that consist of logic gates, e.g., AND and OR gates (see
\cite[Section~2.3]{Raymond95}).  A materialization graph can also be
seen as a Boolean circuit by treating the nodes as AND gates.\\
%
First, the evaluation of a Datalog program is equivalent to the evaluation
of its corresponding Boolean circuit in terms of computational
complexity \cite{DMRT14a}. \\
%
Second, a Boolean circuit can be simulated by PRAM based on an
algorithm called parallel prefix computation \cite{LaFi80a}.  Further,
the complexity of simulation is lower than that of the evaluation of
the Boolean circuit \cite[Lemma~2.4.1]{Raymond95}.\\
%
Third, during the above simulation procedure, a one-to-one
correspondence between each gate in the Boolean circuit and the
processors in PRAM and the constant-time access are established.\\
%
In our case, the one-to-one correspondence between processors and rule
instantiations and the constant-time access to the atoms within the corresponding
rule instantiations can also be
established by PRAM similarly to the above simulation procedure.

Since we focus on the parallel tractability of ontology materialization,
we did not explain the constant time claims in detail.
We use `\emph{the parallel condition}' in Section~3.3 to directly
argue that a one-to-one correspondence between processors
and rule instantiations and the constant-time access to atoms for processors
can be established. We have added the key references and referred the readers
to the details.

@@

Algorithm~$\mathsf{A}_{\text{opt}}$ also works on the parallel condition.
That is each processor $p$ corresponds to a rule instantiation, e.g., $B_1,\ldots,B_n\rightarrow H$,
and processor $p$ can access the locations of nodes $B_1,\ldots,B_n, H$ in constant time.
In order to compute $S_{\!\textit{\tiny rch}}$, processor $p$ should check the derivability
of nodes $B_1,\ldots,B_n, H$. If only node $B_i$ ($i\in\{1,\ldots n\}$) and node $H$
are still not derived, then processor $p$ creates a pair $(B_i, H)$ in $S_{\!\textit{\tiny rch}}$.
The other elements in $S_{\!\textit{\tiny rch}}$ are similarly computed.
On the parallel condition, the above procedure occupies constant time as well as checking
the applicability of a rule instantiation.
We have added the above content to make the claim that computation of $S_\text{rch}$ costs
constant time clearer.
The procedure of updating $G$ is similar to that of Algorithm~$\mathsf{A}_{\text{bsc}}$.
